---
title: "Models with multiple predictors"
subtitle: "<br><br> Environmental Data Science"
author: "[https://enst222.github.io/website](https://enst222.github.io/website)"
output:
  xaringan::moon_reader:
    css: ["../xaringan-themer.css", "../slides.css"]
    lib_dir: libs
    anchor_sections: FALSE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
---

```{r child = "../setup.Rmd"}
```

```{r packages, echo = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(patchwork)
set.seed(1234)
options(dplyr.print_min = 10, dplyr.print_max = 6)
```

class: middle

# The linear model with multiple predictors

---

## Data: Book weight and volume

The `allbacks` data frame gives measurements on the volume and weight of 15 books, some of which are paperback and some of which are hardback

.pull-left[
- Volume - cubic centimetres
- Area - square centimetres
- Weight - grams
]
.pull-right[
.small[
```{r echo=FALSE}
library(DAAG)
as_tibble(allbacks) %>%
  print(n = 15)
```
]
]

.footnote[
.small[
These books are from the bookshelf of J. H. Maindonald at Australian National University.
]
]

---

## Book weight vs. volume

.pull-left[
```{r}
linear_reg() %>%
  set_engine("lm") %>%
  fit(weight ~ volume, data = allbacks) %>%
  tidy()
```
]
.pull-right[
```{r out.width = "75%", echo = FALSE, fig.align = "right"}
ggplot(allbacks, aes(x = volume, y = weight)) +
  geom_point(alpha = 0.7, size = 3)
```
]

---

## Is the effect of volume on weight different for paperback vs. hardcover books?

.pull-left[
```{r}
linear_reg() %>%
  set_engine("lm") %>%
  fit(weight ~ volume, data = allbacks) %>%
  tidy()
```
]
.pull-right[
```{r out.width = "75%", echo = FALSE, fig.align = "right"}
ggplot(allbacks, aes(x = volume, y = weight, color = cover)) +
  geom_point(alpha = 0.7, size = 3) +
  scale_color_manual(values = c("#E48957", "#071381"))
```
]

---

### We can *add* a categorical variable in our model formula to model the effets of an additional variable

.pull-left[
```{r}
linear_reg() %>%
  set_engine("lm") %>%
  fit(weight ~ volume + cover, data = allbacks) %>%
  tidy()
```
]
.pull-right[
```{r out.width = "75%", echo = FALSE, fig.align = "right"}
mod <- linear_reg() %>%
  set_engine("lm") %>%
  fit(weight ~ volume + cover, data = allbacks) %>%
  tidy()

ggplot(allbacks, aes(x = volume, y = weight, color = cover, shape = cover)) +
  geom_point(alpha = 0.7, size = 3) +
  geom_abline(slope = mod$estimate[2], intercept = mod$estimate[1], color = "#E48957") +
  geom_abline(slope = mod$estimate[2], intercept = mod$estimate[1]+mod$estimate[3], color = "#071381") +
  theme(legend.position = "bottom") +
  scale_color_manual(values = c("#E48957", "#071381"))
```
]

Our model fits a line through observations where cover = `hb` and another line through observations where cover = `pb`
---

## Interpretation of estimates

```{r echo=FALSE}
linear_reg() %>%
  set_engine("lm") %>%
  fit(weight ~ volume + cover, data = allbacks) %>%
  tidy()
```

--
- **Intercept (when cover = `hc`):** Hardcover books with 0 volume are expected to weigh 198 grams, on average. (Doesn't make sense in context.)

--

- **Volume (slope):** *All else held constant*, for each additional cubic centimeter books are larger in volume, we would expect the weight to be higher, on average, by 0.718 grams.

--

- **Cover (intercept when cover = `pb`):** *All else held constant*, paperback books weigh, on average, 184 grams less than hardcover books.
--

---
## *Adding* variables = "main effects"

- **Main effect:** The effect of an independent variable on the dependent variable, averaging across the levels of any other independent variables

--

- In other words, the effect of the independent variable doesn't change depending on the value of other independent variables

--

- In our book example, the effect of volume on weight doesn't change depending on whether the book is hardcover or paperback.

--

- So the slopes of the best fit lines are the same for both paperback and hardcover books, but the intercepts are different.

---

### We can *multiply* by a categorical variable in our model formula to model the effets of an additional variable

.pull-left[
```{r}
linear_reg() %>%
  set_engine("lm") %>%
  fit(weight ~ volume * cover, data = allbacks) %>%
  tidy()
```
]
.pull-right[
```{r out.width = "75%", echo = FALSE, fig.align = "right"}
mod <- linear_reg() %>%
  set_engine("lm") %>%
  fit(weight ~ volume * cover, data = allbacks) %>%
  tidy()

ggplot(allbacks, aes(x = volume, y = weight, color = cover, shape = cover)) +
  geom_point(alpha = 0.7, size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  theme(legend.position = "bottom") +
  scale_color_manual(values = c("#E48957", "#071381"))
```
]

Our model fits a line through observations where cover = `hb` and another line through observations where cover = `pb`. **What is different this time?**
---

## Interpretation of estimates

```{r echo=FALSE}
linear_reg() %>%
  set_engine("lm") %>%
  fit(weight ~ volume * cover, data = allbacks) %>%
  tidy()
```

--
- **Intercept (when cover = `hc`):** Hardcover books with 0 volume are expected to weigh 162 grams, on average. (Doesn't make sense in context.)
--

- **Volume (slope when cover = `hc`):** *All else held constant*, for each additional cubic centimeter books are larger in volume, we would expect the weight to be higher, on average, by 0.762 grams.
--

- **Cover (intercept when cover = `pb`):** *All else held constant*, paperback books weigh, on average, 120 grams less than hardcover books.
--

- **volume (slope when cover = `pb`):** *All else held constant*, for each additional cubic centimeter books are larger in volume, we would expect the weight to be higher, on average, by 0.762 - 0.0757 grams (or 0.6863 g).
---
## *Multiplying by* variables = "interaction effects"

- **Interaction effect:** The effect of an independent variable on the dependent variable, as a function of another independent variable

--

- In other words, the effect of the independent variable *does* change depending on the value of other independent variables

--

- In our book example, the effect of volume on weight is different depending on whether the book is hardcover or paperback.

--

- So both the slopes and the intercepts of the best fit lines are different for paperback and hardcover books.

---
class:inverse, middle

# Which model should we use?

---

```{r book-main-int, echo=FALSE, out.width="65%", fig.asp = 0.8}
book_main_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(weight ~ volume + cover, data = allbacks)
book_main_fit_aug <- augment(book_main_fit$fit)

book_int_fit <- linear_reg() %>%
  set_engine("lm") %>%
  fit(weight ~ volume + cover + volume*cover, data = allbacks)
book_int_fit_aug <- augment(book_int_fit$fit)

p_main <- ggplot() +
  geom_point(book_main_fit_aug, 
             mapping = aes(x = volume, y = weight, color = cover, shape = cover), alpha = 0.7) +
  geom_line(book_main_fit_aug, 
            mapping = aes(x = volume, y = .fitted, color = cover)) +
  labs(title = "Main effects, parallel slopes", 
       subtitle = "weight = volume + cover") +
  scale_color_manual(values = c("#E48957", "#071381"))

p_int <- ggplot() +
  geom_point(book_int_fit_aug, 
             mapping = aes(x = volume, y = weight, color = cover, shape = cover), alpha = 0.7) +
  geom_line(book_int_fit_aug, 
            mapping = aes(x = volume, y = .fitted, color = cover)) +
  labs(title = "Interaction effects, not parallel slopes", 
       subtitle = "weight = volume + cover + volume * cover") +
  scale_color_manual(values = c("#E48957", "#071381"))

p_main /
  p_int +
  plot_layout(guides = "collect")
```

---

## In pursuit of Occam's razor

- Occam's Razor states that among competing hypotheses that predict equally well, the one with the fewest assumptions should be selected. 

--

--
- Model selection follows this principle.

--
- We only want to add another variable to the model if the addition of that variable brings something valuable in terms of predictive power to the model.

--
- In other words, we prefer the simplest best model, i.e. **parsimonious** model.

---

## R-squared

- $R^2$ is the percentage of variability in the response variable explained by 
the regression model.

```{r}
glance(book_main_fit)$r.squared
glance(book_int_fit)$r.squared
```

--
- The model with interactions has a higher $R^2$, but only slightly

--
- In addition, using $R^2$ for model selection in models with multiple explanatory variables is not a good idea as $R^2$ increases when **any** variable is added to the model.

---

## Adjusted R-squared

... a (more) objective measure for model selection

- Adjusted $R^2$ doesn't increase if the new variable does not provide any new 
informaton or is completely unrelated, as it applies a penalty for number of 
variables included in the model.

- This makes adjusted $R^2$ a preferable metric for model selection in multiple
regression models.

---

## Comparing models

.pull-left[
```{r}
glance(book_main_fit)$r.squared
glance(book_int_fit)$r.squared
```
]
.pull-right[
```{r}
glance(book_main_fit)$adj.r.squared
glance(book_int_fit)$adj.r.squared
```
]

--

.pull-left-wide[
.small[
- Is R-sq higher for interaction model?
```{r}
glance(book_int_fit)$r.squared > glance(book_main_fit)$r.squared 
```

- Is R-sq adj. higher for interaction model?

```{r}
glance(book_int_fit)$adj.r.squared > glance(book_main_fit)$adj.r.squared
```
]
]
