---
title: "Recoding data"
subtitle: "<br><br> Environmental Data Analysis and Visualization"
author: "[https://mrne222-sp26.github.io/website/](https://mrne222-sp26.github.io/website/)"
format: 
  revealjs:
    theme: simple
    width: 1500
    margin: 0.05
    scrollable: true
slide-number: c/t
from: markdown+emoji
editor: 
  markdown: 
    wrap: 72
execute:
  echo: true
---

## What is "recoding" data?

You've already done some recoding!

- Convert to factor data and recode/relevel using the forcats functions
- Tell R how to handle missing values

## What is "recoding" data?

You might also:

- Rename the levels of a categorical variable (e.g., convert species abbreviations to full species names or vice-versa)
- Create categories based on continuous data values (e.g., convert nutrient concentrations to "low", "medium", and "high")
- Combine multiple categories into fewer more meaningful categories
- Convert text data into numeric codes (e.g., presence/absence to 1/0)


## Recoding data: *adding* a new variable

```{r  echo=FALSE}
library(tidyverse)
```

*Adding* a variable keeps the original column and adds a new column containing the recoded data.

::: {.columns}
::: {.column width="50%"}

```{r}
data <- tibble(types = c("puppy", "dog", "kitten", "cat"))

data
```

:::
::: {.column width="50%"}

```{r}
data |>
  mutate(age = case_match(types, c("puppy", "kitten") ~ "juvenile",
                        c("dog", "cat") ~ "adult"))
```
::: 
:::

## Recoding data: *overwriting* and existing variable

*Overwriting* a variable: the original data are overwritten with the recoded data.

::: {.columns}
::: {.column width="50%"}

```{r}
data <- tibble(types = c("puppy", "dog", "kitten", "cat"))

data
```

:::
::: {.column width="50%"}
```{r}
data |>
  mutate(types = case_match(types, c("puppy", "kitten") ~ "juvenile",
                        c("dog", "cat") ~ "adult"))
```
::: 
:::


## There are a LOT of different ways to recode and transform your data in R. 
<br>
We will cover a few of them, but Google is your friend to figure out how to recode in your own unique use cases.

# Recoding functions

## Examples using the `iris` dataset
::: columns
::: {.column width = "50%" .fragment}
First load packages:
```{r packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
```
:::
::: {.column width = "50%" .fragment}
Clean the data names (`iris` is a built-in dataset that comes with R)
```{r}
iris <- as_tibble(iris) |>
  clean_names()

iris
```
::: 
:::

## `case_match()` to recode values manually

How to use in `mutate()`:

```{r eval=FALSE}
data |> mutate(new_variable, case_match(original_variable,
                                "name1" ~ "new_name1",
                                "name2" ~ "new_name2"))
```

                                                    

## Use `case_match()` to recode species so that setosa = "a", versicolor = "b", and virginica = "c"

::: columns
::: {.column width = "40%" .fragment}
```{r}
distinct(iris, species)
```
:::
::: {.column width = "60%" .fragment}
```{r}
#| code-line-numbers: "2,3,4"
iris |>
  mutate(species = case_match(species, "setosa" ~ "a", 
                              "versicolor" ~ "b", 
                              "virginica" ~ "c")) |> 
  distinct(species)
```
::: 
:::

## Use the `.default` argument if you want to keep some values the same

```{r}
#| code-line-numbers: "3"

iris |>
  mutate(species = case_match(species, "setosa" ~ "a",
                              .default = species)) |> 
  distinct(species)
```

## `if_else()` to recode values based on conditional statements
<br>
Create a `new_variable` based on a `numeric_variable` so that if `numeric_variable` is greater than 10, `new_variable` = "large"; otherwise `new_variable` = "small".
<br>
```{r eval=FALSE}
data |>
  mutate(new_variable = if_else(numeric_variable > 10, "large", "small"))
```


## Use `mutate()` and `if_else()` to replace values based on conditional statement 

**If** `petal_length` is less than three, recode `petal_length` to "short". **Otherwise** ("*else*") recode `petal_length` to "long".
```{r}
iris
```

## Use `mutate()` and `if_else()` to replace values based on certain conditions

**If** `petal_length` is less than three, recode `petal_length` to "short". **Otherwise** ("*else*") recode `petal_length` to "long".
```{r}
#| code-line-numbers: "2"

iris |>
  mutate(petal_length = if_else(petal_length < 3, "short", "long")) 
```

## Create a new variable if you want to preserve the original values

```{r}
#| code-line-numbers: "2"

iris |>
  mutate(petal_category = if_else(petal_length < 3, "short", "long"))
```

## Use `case_when()` for more complicated conditional statements

If `petal_length` is less than or equal to 2, recode `petal_length` to "small". If `petal_length` is greater than 2 and less than or equal to 5, recode to "medium". If `petal_length` is greater than 5, recode to "large".
```{r}
#| code-line-numbers: "2,3,4"

iris |>
  mutate(petal_category = case_when(petal_length <= 2 ~ "small", 
                                    petal_length > 2 & sepal_length <= 5 ~ "medium", 
                                    petal_length > 5 ~ "large")) |> 
  distinct(petal_category)
```

## Use `TRUE` to index values that were not already defined

For all other values of `petal_length`, recode to "other".
```{r}
#| code-line-numbers: "5"

iris |>
  mutate(petal_category = case_when(petal_length <= 2 ~ "small", 
                                    petal_length > 2 & sepal_length <= 5 ~ "medium", 
                                    petal_length > 5 ~ "large",
                                    TRUE ~ "other")) |> 
  distinct(petal_category)
```

## Use joins to recode when you have **many** different values that need to be recoded

<b> For example, you your research collaborator may have used different sample ids than you and you need to recode your data in order for consistency. 

::: {.columns}
::: {.column width="50%"}
```{r}
their_data <- tibble(
  site = c("piney_point", "smcm", "st_inigoes"),
  temp_c = c(30, 24, 27))

their_data
```
:::
::: {.column width="50%"}
```{r}
my_data <- tibble(
  station = c("station1", "station2", "station3"),
  chl_a_ug_l = c(15, 12, 13))

my_data
```
::: 
:::

## Use joins to recode when you have **many** different values that need to be recoded

::: {.columns}
::: {.column width="50%"}
First create a data frame containing their site ids and your corresponding site ids.
```{r}
ids <- tibble(
  site = c("piney_point", "smcm", "st_inigoes"),
  station = c("station1", "station2", "station3"))
```

:::
::: {.column width="50%"}
Use `left_join()` to join the ids dataframe to your dataframe.
```{r}
my_data |>
  left_join(ids)
```
::: 
:::

## Use joins to recode when you have **many** different values that need to be recoded

::: {.columns}
::: {.column width="50%"}
You can keep the resulting dataframe as is:
```{r}
my_data |>
  left_join(ids)
```

:::
::: {.column width="50%"}
Or remove your station names so your sample ids matches your collaborator's ids. 
```{r}
my_data |>
  left_join(ids) |>
  select(-station)
```
::: 
:::

# An example, with some data wrangling

## Calculate number of category 3, 4, and 5 storms each decade and plot the results. 
::: columns
::: {.column width = "50%"}
The data:
```{r}
storms |>
  select(name, year, month, day, hour, category)
```
:::
::: {.column width = "50%"}
The process:

::: {.incremental}
- filter to remove years from decades that have incomplete data
- filter so storm category is greater than or equal to 3
- create new variable (decade)
- calculate total number of storms for each category and decade
:::
::: 
:::


## Filter so year is between 1980 and 2020 and category >= 3 

We need to remove years from decades where data is incomplete.
```{r}
#| code-line-numbers: "3,4,5"

storms |>
  select(name, year, month, day, hour, category) |> 
  filter(year >= 1980, 
         year < 2020, 
         category >= 3) 
```

## Group by year and storm name and calculate the max category reached for each storm and year
::: {.panel-tabset}

## Code
```{r, eval = FALSE}
#| code-line-numbers: "6,7"
#| 
storms |>
  select(name, year, month, day, hour, category) |> 
  filter(year >= 1980, 
         year < 2020, 
         category > 3) |>
  group_by(name, year) |> 
  summarize(maxcat = max(category))
```

## Output
```{r, echo = FALSE}
storms |>
  select(name, year, month, day, hour, category) |> 
  filter(year >= 1980, 
         year < 2020, 
         category > 3) |>
  group_by(name, year) |> 
  summarize(maxcat = max(category)) 
```
:::

## Create a `decade` variable using `mutate()` and `case_when()` based on the `year` variable

::: {.panel-tabset}

## Code
```{r, eval = FALSE}
#| code-line-numbers: "8,9,10,11"
storms |>
  select(name, year, month, day, hour, category) |>
  filter(year >= 1980, 
         year < 2020, 
         category > 3) |>
  group_by(name, year) |>
  summarize(maxcat = max(category)) |>
  mutate(decade = case_when(year %in% 1980:1989 ~ "1980s", 
                            year %in% 1990:1999 ~ "1990s",
                            year %in% 2000:2009 ~ "2000s", 
                            year %in% 2010:2019 ~ "2010s")) 
```

## Output
```{r, echo = FALSE}
storms |>
  select(name, year, month, day, hour, category) |>
  filter(year >= 1980, 
         year < 2020, 
         category > 3) |>
  group_by(name, year) |>
  summarize(maxcat = max(category)) |>
  mutate(decade = case_when(year %in% 1980:1989 ~ "1980s", 
                            year %in% 1990:1999 ~ "1990s", 
                            year %in% 2000:2009 ~ "2000s", 
                            year %in% 2010:2019 ~ "2010s")) 
```
:::

## Group by the new decade variable and count the number of storms in each decade

::: {.panel-tabset}

## Code
```{r, eval = FALSE}
#| code-line-numbers: "12,13"
storms |>
  select(name, year, month, day, hour, category) |> 
  filter(year >= 1980, 
         year < 2020, 
         category > 3) |>
  group_by(name, year) |>
  summarize(maxcat = max(category)) |>
  mutate(decade = case_when(year %in% 1980:1989 ~ "1980s",
                            year %in% 1990:1999 ~ "1990s",
                            year %in% 2000:2009 ~ "2000s",
                            year %in% 2010:2019 ~ "2010s")) |>
  group_by(decade) |> 
  summarize(n = n()) 
```

## Output
```{r, echo = FALSE}
storms |>
  select(name, year, month, day, hour, category) |> 
  filter(year >= 1980, 
         year < 2020, 
         category > 3) |>
  group_by(name, year) |>
  summarize(maxcat = max(category)) |>
  mutate(decade = case_when(year %in% 1980:1989 ~ "1980s",
                            year %in% 1990:1999 ~ "1990s",
                            year %in% 2000:2009 ~ "2000s",
                            year %in% 2010:2019 ~ "2010s")) |>
  group_by(decade) |> 
  summarize(n = n()) 
```
:::

## and then plot!

::: {.panel-tabset}

## Code
```{r "storm-plot", fig.show="hide"}
#| code-line-numbers: "13,14,15,16,17"
storms |>
  filter(year >= 1980,
         year < 2020,
         category > 3) |>
  group_by(name, year) |>
  summarize(maxcat = max(category)) |>
  mutate(decade = case_when(year %in% 1980:1989 ~ "1980s",
                            year %in% 1990:1999 ~ "1990s",
                            year %in% 2000:2009 ~ "2000s",
                            year %in% 2010:2019 ~ "2010s")) |>
  group_by(decade) |>
  summarize(n = n()) |>
  ggplot(aes(x = decade, y = n)) + 
  geom_bar(stat = "identity") + 
  labs(x = "", y = "Number of category 3+ hurricanes") + 
  theme_light() 
```

## Plot
```{r ref.label="storm-plot", echo = FALSE, out.width="60%"}
```
:::